Abstract

### 2025-10-08 output/Response_to_action_editor_reviewers-2.Rmd

Finally, the sentence that was found difficult to understand in the abstract (p. 1, lines 5-7) has been rewritten as follows:

>  [...]
We apply Reckase's maximum discrimination point method to define them in an
arbitrary algebraic basis.
Then, we define such basis to be a
geometrical representation of the measured construct. [...]

###### 2) I would like to see the graphics improved as suggested by Thissen's Review 2.  Labeling of axes is key for clarity.  And providing contour plots of the factor score ( latent variable) distributions will also help the reader’s understanding, especially for those readers coming from the IRT perspective.

We have made the changes proposed, updated the figure caption accordingly, and added an appropriate footer.
These changes have required a rather high effort to make sure they represent the algebra adequately, as well as some edits to improve the clarity of the plots and justify the computations.
Please see our detailed responses to Thissen's comments below.

@@

<!-- # TODO: Discuss complex items, and the proper representation in the test space. -->
<!-- # TODO: Discuss "latent space" transformation is an artifact, not something to "explore". -->

# Responses to Reviewer 4:

###### It is always best, many would say required, to label the axes. The lack of axis labels in the graphics of Figure 1 make the rest of these comments difficult to type, so for convenience I imagine the axes are labeled conventionally x and y in the left panel, and x and y' in the right panel, indicating that y' is a rotated representation of y. (In the ms., it would be best to tie the labeling notation to the algebra used.)

We had overlooked labelling the axes and have done so in this new version;
we thank the reviewer once again for pointing this out.
$`r TRAIT_COMP_1`$/$`r TRAIT_COMP_2`$ have been used as labels for simplicity
(both in the code for plotting them, and for the reader's understanding),
but note that they represent the coordinates of
a latent trait point transformed to the test space,
that is, $`r TRAIT_COORD_1`$/$`r TRAIT_COORD_2`$.
However, there is a direct correspondence between the algebraic components and
the coordinates in either space, so this labeling should be clear enough,
while relying only upon the algebraic interpretation of the model,
thus avoiding the geometrical details,
which are only necessary for deriving the results.
This choice of labeling is explained in the text in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->:

> (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
> for simplicity, but note that they represent the coordinates of
> a latent trait point transformed to the test space,
> that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)

@@

We have added the contour plots to the left panel of Figure 1 (panel (a) in the new version), as suggested.

###### Then one could make another version of the left panel with contours for the rho=0.5 population distribution. Comparison of that with the previously-described version would show that the arrows representing discrimination do not have the same numerical meaning with respect to the variance (standard deviation) of the population distribution *in the direction of the arrows*, which is part of the point of all of this.

###### Two graphics with orthogonal axes but different population distributions are the way orthogonal vs. oblique solutions are represented in MIRT analyses. The ms. misses an opportunity to clarify that. Different population distributions on orthogonal axes are the way MIRT folks have thought about factor correlation; factor analysts have thought about 'rotation' and oblique axes (the left panel of Figure 1); it's two sides of the same coin.

This is indeed the message we want to convey with our manuscript, and are thankful to the reviewer for the comments.
We have taken into account the suggestion and added the contours plots accordingly.
This is represented as panel (b) in the new version.
The text also explains the different interpretation of the coordinates relative to the distribution now.
Please see the new paragraphs added in third and fourth place in section "Graphical Representation Example" (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):

> In panel (b), the item coordinates are computed from
> the agnostic version of the parameters.
> These coordinates are the same ones as in the first case, but
> their relative position with respect to the latent trait distribution is not
> ---This is the effect of disregarding the covariance structure.
> Panel (c) shows the items with their coordinates computed using
> the correlation-based version of the parameters. [...]
>
> Comparing it to panel (b), we can see that the item coordinates are in general
> not the same as the ones computed with the agnostic version of the parameters;
> this effect is especially salient in items 2 and 3,
> which are close to the 90% contour in the agnostic version,
> but are almost along the 50% contour in the correlation-based one. [...]

@@

###### That said, one could also add elliptical contours representing the population distribution to the left panel of Figure 1. That would be a little complicated to compute; it's a "squeezed in" ellipsoid due to the rotation of the y axis to y'. but when compared to a version of the right panel with ellipses for a correlation of 0.5, the arrows would represent the same numerical values of discrimination relative to the standard deviation of the population in the arrows' directions.

As suggested, we have added these elliptical contours.
The computation is also explained in the text; to clarify it, we have had to add some additional results in previous parts of the manuscript.
The changes we have made in this new version are the following:

- The second paragraph of section "Relationship Between Test Space and Latent Space" explains how coordinates can be converted from the latent trait to the test space, and vice versa.
For explaining this, we have needed to add and reference two new equations.
This paragraph (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads as follows:

> [...]
> First, the transformation to the orthonormal bases of the two spaces allows
> mapping the $`r MODEL_ACRONYM`$ parameters interchangeably in the two spaces,
> by projecting one space onto the other;
> that is, we can set
> 
> \begin{equation}
>   `r TRAIT_COORDS_ORTH_SPACES_EQ` \t (23)
> \end{equation}
> 
> 
> for any point, $`r TRAIT_VECTOR`$ or $`r DISCR_VECTOR`$.
> This allows representing the coordinates of the latent space vectors
> in the test space and, conversely,
> the coordinates of the discrimination vectors in the latent space:
> By Equations (2), (5), and (23)
> any coordinate vector $`r TRAIT_COORDS`$ in $`r LS_BASIS`$
> can be expressed in $`r TEST_SPACE_BASIS`$ as
> 
> \begin{equation}
>   `r TRAIT_COORDS_TS_EQ`, \t (24)
> \end{equation}
>   
> and vice versa. [...]

- The second paragraph in section "Graphical Representation" now clarifies better that the representation in the test space must use its basis instead of the canonical basis, to make the plot be a faithful representation of the actual directions of the items with respect to the latent trait dimensions.
This paragraph now reads (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):

> [...] Coordinates in a rectangular Cartesian system are commonly assumed
> to be simply represented in the canonical basis.
> To represent an item in the test space properly, however,
> we must use its basis $`r TEST_SPACE_BASIS`$,
> which implies pre-multiplying its coordinates
> by $`r TRANSFORM_MATRIX_TRANSP_INV`$
> to find the equivalent coordinates in the canonical basis. [...]

- The section "Graphical Representation Example" has been expanded to explain better both Table 1 and Figure 1.
Starting in the first paragraph (p. <!-- # TODO: Complete -->, l. <!-- # TODO: Complete -->), this section now reads:

> Their parameters are shown in Table 1,
> both in the original `r MODEL_ACRONYM` metric and
> in the multidimensional metric,
> in two different cases with different latent covariance structures.
> In the first case (columns $`r CORR_ORTH_OUT`$),
> the two latent dimensions are independent;
> in the second one (columns $`r CORR_ORTH_OUT`$),
> the correlation between the dimensions is .5.
> The same `r MODEL_ACRONYM` parameters are used in both cases,
> but their (correlation-based) multidimensional parameters change when different
> correlations are used.
> 
> Panel (a) of Figure 1 shows
> the item representation in the uncorrelated case,
> while the correlated case is shown in panels (b) to (d)
> under different conditions.
> In all the four panels, the contours of a bivariate standard normal density
> with the corresponding correlation are plot along with the items
> (from outer to inner, the contours are the `r contour_breaks_out` of
> the density at its mode).
> Note that the stretched ellipses in panel (c) are the result of transforming
> the distribution in the latent space to the test space,
> according to Equation (24).
> This implies that the covariance matrix of the distribution projected into
> the test space is transformed inversely as in the latent space, that is,
> $`r COV_MATRIX_TRANSF_TS_EQ`$.
> 
> In panel (b), the item coordinates are computed from
> the agnostic version of the parameters.
> These coordinates are the same ones as in the first case, but
> their relative position with respect to the latent trait distribution is not
> ---This is the effect of disregarding the covariance structure.
> Panel (c) shows the items with their coordinates computed using
> the correlation-based version of the parameters. [...]
> 
> Comparing it to panel (b), we can see that the item coordinates are in general
> not the same as the ones computed with the agnostic version of the parameters;
> this effect is especially salient in items 2 and 3,
> which are close to the 90% contour in the agnostic version,
> but are almost along the 50% contour in the correlation-based one. [...]
> 
> A comparison of panels (a) and (c) allows understanding the effect of
> a positive correlation on the $`r MDISC_CORR_PARAM`$
> (and consequently the $`r MIL_CORR_PARAM`$) value:

@@

Finally, we have added a fourth panel (panel (d)) that illustrates how the items would be plot if we omitted using the proper test space basis (and just using the canonical basis instead).
Incidentally, this also serves the purpose of allowing an easier comparison of the coordinates computed from the two versions (the agnostic, and the correlation-based one) of the multidimensional parameters.
We have briefly explained this, adding in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->:

> Finally, panel (d) illustrates the effect of plotting the coordinates directly
> on the canonical basis; the relative positions and distances are correctly
> represented (which allow a more direct comparison with panel (b)),
> but the angles of the items with the axes are distorted:
> For example, item 1 forms 45-degree angles with both axes,
> instead of the 30-degree ones shown in Table 1.

### 2025-10-08 output/Response_to_action_editor_reviewers-2.Rmd

We have made the changes proposed, updated the figure caption accordingly, and added an appropriate footer.
These changes have required a rather high effort to make sure they represent the algebra adequately, as well as some edits to improve the clarity of the plots and justify the computations.
Please see our detailed responses to Thissen's comments below.

###### 3) I would like to see additional comments in their closing paragraph on how testing practitioners should use or what insights can be obtained by reviewing the different representations of the latent ability space and the item space. I see the mention of multigroup equating/linking, but I would like to see more discussion about how the test results and subsequent uses could be affected.

<!-- # TODO: Discuss complex items, and the proper representation in the test space. -->
<!-- # TODO: Discuss "latent space" transformation is an artifact, not something to "explore". -->

In addition to the previous changes, we have made a minor addition in the "Covariance-based version" section, to refer to the transformation that turns the latent trait distribution into a standard normal by its usual name in the literature, i.e., a whitening transformation.
The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:

> Therefore, an orthonormal basis should represent independent,
> standard coordinates.
> Formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
> $`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$,
> which implies that Equation \@ref(eq:change-basis) actually represents
> a *whitening transformation* [@fukunaga_introduction_2013].
> Given this, $`r COV_MATRIX_TRANSF_EQ`$, and thus

(Please also note the update in the title page to acknowledge the colleague who brought this previous literature to our attention, as well as the valuable contributions of the reviewers.)

### 2025-10-09 output/Main_text.Rmd
Reorder "Graphical Representation Example" content

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version, however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
For graphical representation purposes,
the correlation-based version will thus be preferred.

@@

Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions
is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
(The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
for simplicity, but note that they represent the coordinates of
a latent trait point transformed to the test space,
that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
Their parameters are shown in Table \@ref(tab:example-items-table-out),
both in the original `r MODEL_ACRONYM` metric and
in the multidimensional metric,
in two different cases with different latent covariance structures.
In the first case (columns $`r CORR_ORTH_OUT`$),
the two latent dimensions are independent;
in the second one (columns $`r CORR_ORTH_OUT`$),
the correlation between the dimensions is .5.
The same `r MODEL_ACRONYM` parameters are used in both cases,
but their (correlation-based) multidimensional parameters change when different
correlations are used.

Panel (a) of Figure \@ref(fig:item-plot-out) shows
the item representation in the uncorrelated case,
while the correlated case is shown in panels (b) to (d)
under different conditions.
In all the four panels, the contours of a bivariate standard normal density
with the corresponding correlation are plot along with the items
(from outer to inner, the contours are the `r contour_breaks_out` of
the density at its mode).
Note that the stretched ellipses in panel (c) are the result of transforming
the distribution in the latent space to the test space,
according to Equation \@ref(eq:trait-coords-spaces-transform).
This implies that the covariance matrix of the distribution projected into
the test space is transformed inversely as in the latent space, that is,
$`r COV_MATRIX_TRANSF_TS_EQ`$.

In panel (b), the item coordinates are computed from
the agnostic version of the parameters.
These coordinates are the same ones as in the first case, but
their relative position with respect to the latent trait distribution is not
---This is the effect of disregarding the covariance structure.
Panel (c) shows the items with their coordinates computed using
the correlation-based version of the parameters.
To keep the horizontal axis invariant,
we have defined $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
it only rotates the vertical axis [@harman_modern_1976]:

@@

Comparing it to panel (b), we can see that the item coordinates are in general
not the same as the ones computed with the agnostic version of the parameters;
this effect is especially salient in items 2 and 3,
which are close to the 90% contour in the agnostic version,
but are almost along the 50% contour in the correlation-based one.
Finally, panel (d) illustrates the effect of plotting the coordinates directly
on the canonical basis; the relative positions and distances are correctly
represented (which allow a more direct comparison with panel (b)),
but the angles of the items with the axes are distorted:
For example, item 1 forms 45-degree angles with both axes,
instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).

>>>>

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based version,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version, however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
For graphical representation purposes,
the correlation-based version will thus be preferred.

@@

Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions
is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items with two different latent trait distributions.
(The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
for simplicity, but note that they represent the coordinates of
a latent trait point transformed to the test space,
that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
Their parameters are shown in Table \@ref(tab:example-items-table-out),
both in the original `r MODEL_ACRONYM` metric and
in the multidimensional metric under the two cases,
which differ in the latent trait correlation.
In the first case (columns $`r CORR_ORTH_OUT`$),
the two latent dimensions are independent;
in the second one ($`r CORR_OBL_OUT`$),
the correlation between them is .5.
The same `r MODEL_ACRONYM` parameters are used in both cases,
but their (correlation-based) multidimensional parameters change
with the correlation.
If we used the agnostic version instead, the multidimensional parameters,
given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.

Panel (a) of Figure \@ref(fig:item-plot-out) shows
the item representation in the uncorrelated case,
while the $`r CORR_OBL_OUT`$ case is shown in panels (b) to (d)
under different conditions:
with item coordinates computed from the agnostic version of the parameters
(panel (b)), or from the correlation-based version,
either using an appropriate oblique basis (panel (c)),
or plotting them directly in the canonical basis (panel (d)).
Keeping the horizontal axis invariant in panel (c)
allows for a better comparison with the other plots;
therefore, we have chosen a $`r TRANSFORM_MATRIX_TRANSP_INV`$ that
only rotates the vertical axis [@harman_modern_1976]:

@@

In all four panels, the contours of a bivariate standard normal density
with the corresponding correlation are plot along with the items
(from outer to inner, the contours are the `r contour_breaks_out` of
the density at its mode).
The stretched ellipses in panel (c) are the result of transforming
the distribution in the latent space to the test space,
according to Equation \@ref(eq:trait-coords-spaces-transform).
This implies that the covariance matrix of the distribution projected into
the test space is transformed inversely as in the latent space, i.e.,
$`r COV_MATRIX_TRANSF_TS_EQ`$.

Panels (a) and (b) show that,
although the agnostic coordinates are the same in both cases,
their relative position with respect to the latent trait distribution is not
---This is the effect of disregarding the covariance structure.
The correlation-based coordinates, plot in panels (c) and (d),
are also different in general from their agnostic counterpart (panel (b)).
This difference is especially salient in the location of items 2 and 3,
which are close to the 90% contour in the agnostic version,
but are almost along the 50% contour in the correlation-based one.
Panel (d) also illustrates the effect of plotting the coordinates directly
on the canonical basis;
the item locations and relative positions to the distribution
are correctly represented (allowing a more direct comparison with panel (b)),
but their angles with the axes and discrimination parameters are distorted:
For example, item 1 forms 45-degree angles with both axes,
instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out);
items 4 and 5 seem to be equally discriminating,
when Table \@ref(tab:example-items-table-out) panel (c) show they are not.
